---
title: "spaCBA"
author: "Sander Wuyts"
date: "31 August 2018"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)

files <- str_c("out/spaCBA/", list.files("out/spaCBA/",  pattern = ".tsv")) 
names(files) <- files

spaCBA <- map_df(files, read_tsv,
                col_names = c("target_name","gene","percentage_identity","length","query_coverage"),
                .id = "filename") %>%
  separate(filename, "/", into = c("deleteme", "deleteme2", "file"), fill = "right") %>%
  separate(file, "\\.", into = c("Assembly", "deleteme3"), fill = "right") %>%
  select(-deleteme, -deleteme2, -deleteme3)

# Read in clade info
clades <- read_tsv("~/serverUA/sander/mudan/07_ANI/genomeTableWithClades.tsv")%>%
  select(Assembly, species, cladeNumber_split)


# Add clade number
spaCBA <- spaCBA %>%
  left_join(clades) 
```

General hits

```{r}
ggplot(spaCBA, aes(x = query_coverage, y = percentage_identity)) +
  geom_point() +
  facet_wrap(~target_name)
```

There does not seem to be a decent hit. However, there's one with high query coverage of spaC; which one is that?

```{r}
spaCBA %>% 
  filter(query_coverage > 80,
         percentage_identity > 50)


```

The strain is called Lactobacillus plantarum LZ227, which has been sequenced with PacBio. It seems that this spaC and the srtC gene are laying on a plasmid. 
More about the strain here:
https://www.sciencedirect.com/science/article/pii/S0168165616314304?via%3Dihub#fig0015

It seems that it's also able to produce vitamins


Nothing more interesting to report.



