---
title: "ANI"
author: "Sander Wuyts"
date: "27 August 2018"
output: html_document
---


# Add clades to genomeTable based on ANI

```{r}
library(tidyverse)
library(ggtree)
library(ape)
library(stringr)

pin_qc = "~/serverUA/sander/mudan/03_quality_control/"
pin_ani = "~/serverUA/sander/mudan/07_ANI/pyani_output/ANIb/"
pout = "~/serverUA/sander/mudan/07_ANI/"

# Following files are required as input:
# - genomeTable.tsv
# - genomeTableFull.tsv
# - ANIb_percentage_identity.tab

# read and parse tables
p_genomeTable = paste(pin_qc, "genomeTable.tsv", sep = '')
p_genomeTableFull = paste(pin_qc, "genomeTable_full.tsv", sep = '')
p_ANItable = paste(pin_ani, "ANIb_percentage_identity.tab", sep = '')
genomeTable = read_tsv(p_genomeTable, col_names = c("Assembly", "species")) 
genomeTableFull = read_tsv(p_genomeTableFull, col_names = T) 
ANIb = read.table(p_ANItable)

# extract clades from ANI table
clust = hclust(as.dist(1-ANIb))
cladeNumbers = cutree(clust, 7) %>% 
  tbl_df() %>% 
  rownames_to_column("Assembly") %>% 
  rename(cladeNumber = value)
genomeTable <- genomeTable %>% left_join(cladeNumbers)

# add subspecies name
genomeTableFull$subspecies <- str_extract(genomeTableFull$organism_name,"(subsp\\.\\s)([a-z]+)")
genomeTable <- genomeTable %>%
  left_join(genomeTableFull %>% select(Assembly, subspecies, infraspecific_name)) %>%
  mutate(strain = str_split_fixed(infraspecific_name, pattern = "=", n = 2)[,2]) %>%
  select(-infraspecific_name)
genomeTable[is.na(genomeTable$subspecies),"subspecies"] <- ""
```

```{r}
genomeTable %>%
  group_by(cladeNumber) %>%
  count(species) %>% 
  summarise(Relabun = max(n)/sum(n), Species = species[which.max(n)], Count = max(n)) 
```

Rename cladenumbers based on how it is written in publication

```{r}
genomeTable <- genomeTable %>%
  mutate(cladeNumber = paste(cladeNumber, "X", sep="")) %>%
  mutate(cladeNumber = recode(cladeNumber, `1X` = '7', 
                              `2X` = '1',
                              `3X` = '5',
                              `4X` = '2',
                              `5X` = '6',
                              `6X` = '4',
                              `7X` = '3')) %>%
  mutate(cladeNumber = str_c("clade", cladeNumber, sep = ''))
```


```{R}
# write genome table with clade names
setwd(pout)
write.table(genomeTable, file="genomeTableWithClades.tsv",
            quote=F, col.names=T, row.names=F, sep="\t")
```


# Read in tree

```{R}
tree <- read.tree('~/serverUA/sander/mudan/06_phylogenetic_tree/raxml_out/RAxML_bipartitions.mudan')

# Root tree
tree_r <- ape::root(tree, "outgroup")
# Remove outgroup
tree_r_wooutroup <- drop.tip(tree_r, "outgroup")

# Test ggtree
p <- ggtree(tree_r_wooutroup) +  geom_tree()
p

# Get the label order
d <- fortify(tree_r_wooutroup)
dd <- subset(d, isTip)
order <- dd$label[order(dd$y,decreasing=T)] 
```


# ANIb
Read in data and reorder dataframes
```{r}
ANIb <- read.table("~/serverUA/sander/mudan/07_ANI/pyani_output/ANIb/ANIb_percentage_identity.tab")

# Fix . to - 
colnames(ANIb) <- gsub(".","-",colnames(ANIb),fixed=T) %>%
  str_replace_all("GCA_","GCA-")

rownames(ANIb) <- rownames(ANIb) %>%
  str_replace_all("GCA_","GCA-")

ANIb <- ANIb[order,order]

ANIbplot <- gheatmap(p, ANIb,width=15,colnames=F, low="yellow", high="red",color=NA,offset=0.1) + 
  ggtitle("ANIb") + scale_fill_gradient(high="#132B43",low="#56B1F7")
ANIbplot
```



# TETRA

```{r}
TETRA <- read.table("~/serverUA/sander/mudan/07_ANI/pyani_output/TETRA/TETRA_correlations.tab")

# Fix . to - 
colnames(TETRA) <- gsub(".","-",colnames(TETRA),fixed=T) %>%
  str_replace_all("GCA_","GCA-")

rownames(TETRA) <- rownames(TETRA) %>%
  str_replace_all("GCA_","GCA-")


TETRA <- TETRA[order,order]

TETRAplot <- gheatmap(p, TETRA,width=15,colnames=F, low="yellow", high="red",color=NA, offset=0.1)  + 
  ggtitle("TETRA") + scale_fill_gradient(high="#132B43",low="#56B1F7")
TETRAplot
```

# Combining

```{R}
library(gridExtra)
grid.arrange(ANIbplot,TETRAplot,nrow=2)
plot <- arrangeGrob(ANIbplot,TETRAplot,nrow=2)
ggsave("Realplot.png",dpi=300,plot=plot,height=10)
```

# Statistics

```{r}
ANIb$Genome1 <- row.names(ANIb) %>%
  str_replace_all("GCA-","GCA_")

cladeNumbers <- genomeTable %>% 
  select(cladeNumber, Assembly)

ANI_tidy <- as_tibble(ANIb) %>%
  gather(key = "Genome2", value = "ANI", -Genome1) %>%
  mutate(Genome2 = str_replace(Genome2, "GCA-","GCA_")) %>%
  left_join(cladeNumbers, by = c("Genome1" = "Assembly")) %>%
  rename(cladeGenome1 = cladeNumber) %>%
  left_join(cladeNumbers, by = c("Genome2" = "Assembly")) %>%
  rename(cladeGenome2 = cladeNumber)  %>%
  mutate(comparison = ifelse(cladeGenome1 == cladeGenome2, "Intraclade", "Interclade"))
  
  
ANI_summary <- ANI_tidy %>% 
  filter(cladeGenome1 == cladeGenome2) %>%
  select(- cladeGenome2) 

ANI_summary %>%
  group_by(cladeGenome1) %>%
  summarise(minANI = min(ANI), meanANI = mean(ANI))
```

```{r, fig.height = 8 , fig.width = 8}
cladenames <- as_labeller(
  c("clade1" = "italic(L.~mudanjiangensis)",
    "clade2" = "italic(L.~fabifermentans)",
    "clade3" = "italic(L.~xiangfangensis)",
    "clade4" = "italic(L.~herbarum)",
    "clade5" = "italic(L.~pentosus)",
    "clade6" = "italic(L.~paraplantarum)",
    "clade7" = "italic(L.~plantarum)"),
  label_parsed
)

nmbrGenomes <- length(unique(ANI_summary$Genome1))

ggplot(ANI_tidy, aes(x = ANI, fill = comparison, colour = comparison)) +
  geom_density(aes(y = ..scaled..), alpha = 0.8) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  geom_vline(xintercept = 0.95, colour = "red") +
  facet_wrap(~cladeGenome2, scales = "free_y", nrow = 7, labeller = cladenames)  +
  ylab("Density") +
  theme(legend.title = element_blank(),
                strip.background = element_blank()) +
  ggtitle("Pairwise ANI comparisons")

ggsave("ANI.png", width = 21, height = 29.7, units = 'cm')
```

Alternative visualisation

```{r, fig.height = 8 , fig.width = 6}
ggplot(ANI_tidy, aes(x = comparison,  y = ANI, colour = comparison, group = comparison)) + 
  geom_jitter(alpha= 0.6, width= 0.3) +
  geom_hline(aes(yintercept = 0.95),  colour = "red") +
  scale_color_brewer(palette = "Paired") +
  facet_wrap(~ cladeGenome1, ncol = 3) +
  xlab("") + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

```
Alternative visualisation

```{r, fig.height = 8 , fig.width = 8}
ggplot(ANI_tidy, aes(x = ANI,  y = 1 , colour = comparison, group = comparison)) + 
  geom_jitter(alpha= 0.6, width= 0, size = 2) +
  geom_vline(aes(xintercept = 0.95),  colour = "red", size = 1.5) +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(~ cladeGenome1, nrow = 7, labeller = cladenames) +
  ylab("") +
    theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background = element_blank(),
        legend.title = element_blank()) +
  ggtitle("Pairwise ANI comparisons")

ggsave("ANI_V2.png", width = 21, height = 29.7, units = 'cm')


```

# Split clade 5 in two sub clades

```{r}
clade5 <- genomeTable %>% 
  filter(cladeNumber == "clade5") %>%
  pull(Assembly) %>%
  str_replace("GCA_","GCA-")

ANIb_clade5 <- ANIb[clade5, clade5]

# extract clades from ANI table
clust = hclust(as.dist(1-ANIb_clade5))
clade5 <- cutree(clust, 2) %>% 
  as_data_frame() %>%
  rownames_to_column() %>%
  rename(Assembly = rowname, cladeNumber = value) %>%
  mutate(cladeNumber_split = str_c("clade",if_else(cladeNumber == 1, "5a", "5b")),
         Assembly = str_replace(Assembly, "GCA-", "GCA_")) %>%
  select(-cladeNumber)

genomeTable <- genomeTable %>%
  left_join(clade5) %>%
  mutate(cladeNumber_split = if_else(is.na(cladeNumber_split), cladeNumber, cladeNumber_split))

# Add cladename
genomeTable <- genomeTable %>%
  mutate(cladeName = case_when(cladeNumber_split == "clade1" ~ "L. mudanjiangensis",
                               cladeNumber_split == "clade2" ~ "L. fabifermentans",
                               cladeNumber_split == "clade3" ~ "L. xiangfangensis",
                               cladeNumber_split == "clade4" ~ "L. herbarum",
                               cladeNumber_split == "clade5a" ~ "clade5a",
                               cladeNumber_split == "clade5b" ~ "L. pentosus",
                               cladeNumber_split == "clade6" ~ "L. paraplantarum",
                               cladeNumber_split == "clade7" ~ "L. plantarum")) %>%
  mutate(cladeName = factor(cladeName, levels = c("L. mudanjiangensis",
                                                  "L. fabifermentans",
                                                  "L. xiangfangensis",
                                                  "L. herbarum",
                                                  "clade5a",
                                                  "L. pentosus",
                                                  "L. paraplantarum",
                                                  "L. plantarum")))



write.table(genomeTable, file="genomeTableWithClades.tsv",
            quote=F, col.names=T, row.names=F, sep="\t")
```

Now let's redo the analysis/vis

## Statistics

```{r}
ANIb$Genome1 <- row.names(ANIb) %>%
  str_replace_all("GCA-","GCA_")

cladeNumbers <- genomeTable %>% 
  select(cladeNumber_split, Assembly)

ANI_tidy <- as_tibble(ANIb) %>%
  gather(key = "Genome2", value = "ANI", -Genome1) %>%
  mutate(Genome2 = str_replace(Genome2, "GCA-","GCA_")) %>%
  left_join(cladeNumbers, by = c("Genome1" = "Assembly")) %>%
  rename(cladeGenome1 = cladeNumber_split) %>%
  left_join(cladeNumbers, by = c("Genome2" = "Assembly")) %>%
  rename(cladeGenome2 = cladeNumber_split)  %>%
  mutate(comparison = ifelse(cladeGenome1 == cladeGenome2, "Intraclade", "Interclade"))
  
  
ANI_summary <- ANI_tidy %>% 
  filter(cladeGenome1 == cladeGenome2) %>%
  select(- cladeGenome2) 

ANI_summary %>%
  group_by(cladeGenome1) %>%
  summarise(minANI = min(ANI), meanANI = mean(ANI))
```

```{r, fig.height = 8 , fig.width = 8}
nmbrGenomes <- length(unique(ANI_summary$Genome1))

ggplot(ANI_tidy, aes(x = ANI, fill = comparison, colour = comparison)) +
  geom_density(aes(y = ..scaled..), alpha = 0.8) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  geom_vline(xintercept = 0.95, colour = "red") +
  facet_wrap(~cladeGenome2, scales = "free_y", nrow = 8)  
```

Alternative visualisation

```{r, fig.height = 8 , fig.width = 6}
ggplot(ANI_tidy, aes(x = comparison,  y = ANI, colour = comparison, group = comparison)) + 
  geom_jitter(alpha= 0.6, width= 0.3) +
  geom_hline(aes(yintercept = 0.95),  colour = "red") +
  scale_color_brewer(palette = "Paired") +
  facet_wrap(~ cladeGenome1, ncol = 2) +
  xlab("") + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

```
Alternative visualisation

```{r, fig.height = 8 , fig.width = 8}
cladenames2 <- as_labeller(
  c("clade1" = "italic(L.~mudanjiangensis)",
    "clade2" = "italic(L.~fabifermentans)",
    "clade3" = "italic(L.~xiangfangensis)",
    "clade4" = "italic(L.~herbarum)",
    "clade5b" = "italic(L.~pentosus)",
    "clade5a" = "paste('clade ',5, 'a')",
    "clade6" = "italic(L.~paraplantarum)",
    "clade7" = "italic(L.~plantarum)"),
  label_parsed
)
ggplot(ANI_tidy, aes(x = ANI,  y = 1 , colour = comparison, group = comparison)) + 
  geom_jitter(alpha= 0.6, width= 0, size = 2) +
  geom_vline(aes(xintercept = 0.95),  colour = "red", size = 1.5) +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(~ cladeGenome1, nrow = 8, labeller = cladenames2) +
  ylab("") +
    theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background = element_blank(),
        legend.title = element_blank()) +
    ggtitle("Pairwise ANI comparisons after split")

ggsave("ANI_V2_split.png", width = 21, height = 29.7, units = 'cm')

```
