---
title: "Gene content analysis L. mudan"
author: "Sander Wuyts"
date: "28 September 2018"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
library(vegan)
library(ggpubr)
library(UpSetR)
library(ggplotify)
library(knitr)
```

# Initial processing

## Read in all data

```{r}
# Orthofinder results
gene_pres_abs <- read_tsv("../05_pangenome/Results_Sep03/Orthogroups.GeneCount.csv") %>%
  rename(orthogroup = X1)%>%
  select(-Total) %>%
  gather(key = genome, value = ngenes, -orthogroup) %>%
  filter(! is.na(ngenes) )

# ANI results
genomeTable <- read_tsv("../07_ANI/genomeTableWithClades.tsv") %>%
    mutate(cladeName = case_when(cladeNumber == "clade1" ~ "L. mudanjiangensis",
                               cladeNumber == "clade2" ~ "L. fabifermentans",
                               cladeNumber == "clade3" ~ "L. xiangfangensis",
                               cladeNumber == "clade4" ~ "L. herbarum",
                               cladeNumber_split == "clade5b" ~ "L. pentosus",
                               cladeNumber_split == "clade5a" ~ "clade5a",
                               cladeNumber == "clade6" ~ "L. paraplantarum",
                               cladeNumber == "clade7" ~ "L. plantarum")) %>%
    mutate(cladeName = factor(cladeName, levels = c("L. mudanjiangensis",
                                                  "L. fabifermentans",
                                                  "L. xiangfangensis",
                                                  "L. herbarum",
                                                  "clade5a",
                                                  "L. pentosus",
                                                  "L. paraplantarum",
                                                  "L. plantarum")))  %>%
  rename(genome = Assembly)

# Eggnog results
eggnog <- read_tsv("../08_map_to_eggnog/out/mudan_pan_genome.emapper.annotations", skip = 3) %>%
  rename(orthogroup = `#query_name`,
         functCat = `COG cat`) %>%
  mutate(orthogroup = str_replace(orthogroup, "mudan_", ""))

# functional categories of eggnog
eggnogFunctCat <- read_tsv("in/functCat_descriptions.tsv")

```

## Merge some tables

```{r}
gene_pres_abs <- gene_pres_abs %>%
  left_join(genomeTable)
```

# Summary statistics of Plantarum group and clades

## Table

```{r}
# adapted from casei group script
calculateCladeStatistics = function(gene_pres_abs) {
  
  part1 = 
    gene_pres_abs %>%
    select(genome)%>%
    distinct()  %>%
    summarise(nGenomes = n())
  nGenomesTot = part1[[1]][1]
  
  part2 = gene_pres_abs %>%
    group_by(orthogroup) %>%
    summarise(ngenomes = sum(ngenes != 0)) %>%
    filter(ngenomes != 0) %>%
    summarise(nCoreOrthogroups = sum(ngenomes >= (0.96 * nGenomesTot)), 
              nAccOrthogroups = sum(ngenomes < (0.96 * nGenomesTot)))
  
  part3 = gene_pres_abs %>%
    group_by(genome) %>%
    summarise(nOrthogroups = sum(ngenes != 0), nGenes = sum(ngenes)) %>%
    summarise(avnOrthogroups = mean(nOrthogroups), avnGenes = mean(nGenes),
              sdnOrthogroups = sd(nOrthogroups), sdnGenes = sd(nGenes))
  
  part4  = gene_pres_abs %>%
    select(genome,species) %>%
    distinct() %>%
    count(species) %>%
    summarise(mostAbunSpecies = .$species[which.max(n)], Abundance = .$n[(which.max(n))]/sum(n))
  return(cbind(part1, part2, part3, part4))
  
}

clade_stats <- gene_pres_abs %>%
  list(., 
       filter(., cladeNumber_split == "clade1"),
       filter(., cladeNumber_split == "clade2"),
       filter(., cladeNumber_split == "clade3"),
       filter(., cladeNumber_split == "clade4"),
       filter(., cladeNumber_split == "clade5a"),
       filter(., cladeNumber_split == "clade5b"),
       filter(., cladeNumber_split == "clade6"),
       filter(., cladeNumber_split == "clade7")) %>%
  lapply(FUN = calculateCladeStatistics) %>%
  bind_rows() %>%
  mutate(clade = c("Plantarum group", "clade 1", "clade 2", "clade 3", "clade 4", "clade 5a", "clade 5b", "clade 6", "clade 7")) %>%
  select(clade,everything())

clade_stats
write_tsv(clade_stats, "clade_stats.tsv")
```

## Visualisation


```{r}
plotdata <-clade_stats %>%
  select(clade, nCoreOrthogroups, nAccOrthogroups) %>%
  gather(key = "type", value = "count", -clade) 

ggplot(plotdata, aes(x = clade, y = count)) +
  expand_limits(y=0) +
  geom_col(aes(fill = type)) +
  geom_text(data = clade_stats, aes(x = clade, y = 9000, label = nGenomes)) +
  coord_flip() +
  scale_fill_brewer(palette = "Dark2")  +
  theme_classic() 

```

```{r, fig.width=12, fig.height=6}
plotdata <- gene_pres_abs %>%
  group_by(genome) %>%
  mutate(nOrthogroups = sum(ngenes != 0), nGenes = sum(ngenes)) %>%
  select(genome, cladeName, nOrthogroups, nGenes) %>%
  distinct() %>%
  gather(key = "type", value = "count", -genome, -cladeName)

nGenes <- plotdata %>%
  filter(type == "nGenes") %>%
  ggplot(aes(x = cladeName, y = count, fill = cladeName, group = cladeName)) +
  geom_boxplot(aes(color = cladeName), alpha = 0.20, outlier.shape =  NA) +
  geom_point(aes(col = cladeName), position = position_jitterdodge()) +
  scale_fill_brewer(palette = "Dark2") +
  scale_colour_brewer(palette = "Dark2") +
  xlab("") +
  ylab("Number of genes") +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) 
nGenes
```

northogroups

```{r}
nOrthogroups <- plotdata %>%
  filter(type == "nOrthogroups") %>%
  ggplot(aes(x = cladeName, y = count, fill = cladeName, group = cladeName)) +
  geom_boxplot(aes(color = cladeName), alpha = 0.20, outlier.shape =  NA) +
  geom_point(aes(col = cladeName), position = position_jitterdodge()) +
  scale_fill_brewer(palette = "Dark2") +
  scale_colour_brewer(palette = "Dark2") +
  xlab("") +
  ylab("Number of orthogroups") +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) 
nOrthogroups
```


Genome sizes

```{r}
# read quast report file
report = read.table("../03_quality_control/out/transposed_report.tsv" , header=T, sep="\t", comment.char="", quote="") %>%
  rename(genome = Assembly)
# column names: replace any number of points by one underscore
names(report) = str_replace_all(names(report), "\\.+", "_")
# column names: remove trailing underscore
names(report) = str_replace(names(report), "\\_$", "")
# column names: replace X by N (number of...)
names(report) = str_replace(names(report), "X", "N")

genomesize <- genomeTable %>%
  left_join(report) %>%
  mutate(genomesize = Total_length/1000000) %>%
  select(genome, cladeName, genomesize) %>%
  gather(key = "type", value = "count", -genome, -cladeName)

genomesize_plot <- genomesize %>%
  ggplot(aes(x = cladeName, y = count, fill = cladeName)) +
  geom_boxplot(aes(color = cladeName), alpha = 0.20, outlier.shape =  NA) +
  geom_point(aes(col = cladeName), position = position_jitterdodge()) +
  scale_fill_brewer(palette = "Dark2") +
  scale_colour_brewer(palette = "Dark2") +
  xlab("") +
  ylab("Genome size (Mb)") +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) 
genomesize_plot
```

GC

```{r}
GC <- genomeTable %>%
  left_join(report) %>%
  select(genome, cladeName, GC) %>%
  gather(key = "type", value = "count", -genome, -cladeName)

gc_plot <- GC %>%
  ggplot(aes(x = cladeName, y = count, fill = cladeName)) +
  geom_boxplot(aes(color = cladeName), alpha = 0.20, outlier.shape =  NA) +
  geom_point(aes(col = cladeName), position = position_jitterdodge()) +
  scale_fill_brewer(palette = "Dark2") +
  scale_colour_brewer(palette = "Dark2") +
  xlab("") +
  ylab("GC-content (%)") +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) 
gc_plot
```

```{r, fig.width=8, fig.height=6}
ggarrange(genomesize_plot,
          gc_plot,
          nGenes,
          ncol = 3,
          legend = 'none',
          labels = "AUTO")
```

```{r, fig.width=8, fig.height=6}
plotdata %>%
  bind_rows(genomesize) %>%
  bind_rows(GC) %>%
  ggplot(aes(x = cladeName, y = count, fill = cladeName, group = cladeName)) +
  geom_boxplot(aes(color = cladeName), alpha = 0.20, outlier.shape =  NA) +
  geom_point(aes(col = cladeName), position = position_jitterdodge()) +
  scale_fill_brewer(palette = "Dark2") +
  scale_colour_brewer(palette = "Dark2") +
  xlab("") +
  ylab("") +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  facet_wrap(~type, scales = "free_y")


```


# Functional analysis à la casei group paper

```{r}
make_pcoaTable = function(gene_pres_abs) {
  
  gene_pres_abs %>%
    select(orthogroup, genome, ngenes) %>%
    spread(key = orthogroup, value = ngenes, fill = 0) %>%
    remove_rownames() %>%
    column_to_rownames(var = "genome") %>%
    as.matrix() %>%
    vegdist(method = "bray") %>%
    cmdscale(k = 2) %>%
    as.data.frame() %>%
    rownames_to_column(var = "genome") %>%
    return()
  
}
```

## PCoA for all orthogroups


```{r}
# make table with pcoa coordinates for all genomes 
T_pcoa1 <- make_pcoaTable(gene_pres_abs) %>%
  left_join(genomeTable)


# plot and save
ggplot(T_pcoa1, aes(x = V1, y = V2, col = cladeNumber_split)) +
  geom_point(size= 3 , alpha= 0.8) +
  scale_colour_brewer(palette = "Dark2") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'))
ggsave(filename = "pcoa_overall.png", height = 20, width = 30, units = "cm")
```

## PCoA per functional category
Next, we can do the same, but per functional category. Some important things to keep in mind:

* It is important to realize that several orthogroups are present in multiple functional categories!! So these are represented in multiple sub-plots. 
* We will remove functional categories with less than 20 orthogroups from the figure because there is too little information to "resolve" all points (genomes) in the pcoa plot. 
* The distances between genomes are percentages, not of a number of orthogroups but of a number of genes!! It would of course be possible to change this to orthogroup presence/absence instead of considering the numbers of genes. This would yield a slightly different graph. 
* The large majority of the orthogroups has no known function! 

```{r fig.width=10, fig.height=5}
gene_pres_abs_eggnog <- gene_pres_abs %>%
  left_join(eggnog) %>%
  select(genome, orthogroup, ngenes, functCat, `eggNOG annot`) %>%
  mutate(functCat = ifelse(is.na(functCat), "S", functCat)) %>% # if function is unkown add to category S
  separate(functCat, into = c("temp1","temp2"), sep=", ", fill = "right") %>% # split rows with multiple functCats
  gather(key = "temp", value = "functCat", temp1, temp2) %>%
  select(-temp) %>%
  drop_na(functCat)

# make table with number of orthogroups per functional category
T_nOrthogroupsPerFunctCat <- gene_pres_abs_eggnog %>%
  select(orthogroup, functCat) %>%
  distinct() %>%
  group_by(functCat) %>%
  summarise(n = n()) %>%
  left_join(eggnogFunctCat) %>%
  mutate(functCatDescription = str_c(functCat, " (n = ", n, ")", "\n", str_wrap(description, width= 55)))

# do pcoa per functional category and add genome info
T_pcoa2 = gene_pres_abs_eggnog %>%
  split(f = .$functCat) %>%
  lapply(FUN = function(table) {
    if (nrow(table) <= 5) return(NULL)
    functionalCategory = table$functCat[1]
    make_pcoaTable(table) %>%
      mutate(functCat = functionalCategory) %>%
      mutate(V1 = V1 - (min(V1) + max(V1)) / 2,
             V2 = V2 - (min(V2) + max(V2)) / 2) %>%     
      return()
  }) %>% bind_rows() %>%
  left_join(genomeTable) %>%
  left_join(T_nOrthogroupsPerFunctCat)

# make and save plot! 
ggplot(T_pcoa2, aes(x = V1, y = V2, col = cladeNumber_split)) +
  geom_point(alpha=0.5, size = 2) +
  scale_colour_brewer(palette = "Dark2") +
  facet_wrap(~ functCatDescription, ncol = 5) +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'))
ggsave(filename = "pcoa_perFunctCat.png", height = 25, width = 40, units = "cm")
```

Highlights:

- Clear difference in Energy production in clades 1-4 vs clades 5-7 -> Need to look into this!!!!!
- Similar for category E
- Clade 1 does something different for Transcirption 
- Clade 1 is different in cell wall biogenesis!!!!!!
- Clade 1 is very different regarding cell motility!!!!!!!
- Clade 1 is different in Posttranslational modification
- Clade 1 is different in inorganic transport
- There's one outlier in Q regarding clade2. need to look into this!!!! -> Does this fabifermentans do something different???

## Digging deeper in functional stuff

### Case study: Category N

There's a difference between OGs presence abscence for category N and clade 1. Can we figure out which OGs?

```{r}
cat_N <- gene_pres_abs_eggnog %>%
  filter(functCat == "N") %>%
  filter(ngenes >0)

cat_N %>%
  group_by(genome) %>% 
  summarise(count = sum(ngenes > 0)) %>%
    left_join(genomeTable) %>%
  ggplot(aes(x= cladeNumber_split, y = count, col = cladeNumber_split)) +
  scale_color_brewer(palette = "Dark2") +
  geom_jitter(height = 0) +
  coord_flip()
```

```{r, fig.height= 10}
cat_N %>% 
  left_join(genomeTable) %>%
  ggplot(aes(x=genome, y= ngenes, fill=orthogroup)) + 
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~cladeNumber_split, scales = "free_x", nrow = 8)

```


# Which are unique orthogroups?

```{r}
clade_unique <- gene_pres_abs %>%
  filter(ngenes > 0) %>%
  select(orthogroup, cladeNumber_split) %>%
  distinct() %>%
  add_count(orthogroup) %>%
  filter(n == 1) %>%
  select(-n) %>%
  left_join(eggnog %>% select(orthogroup, functCat, `eggNOG annot`))

clade_unique %>%
  group_by(cladeNumber_split) %>%
  summarise(count = n())
```



```{r, fig.height= 10, fig.width = 8}
clade_unique %>%
  mutate(functCat = ifelse(is.na(functCat), "S", functCat)) %>% # if function is unkown add to category S
  separate(functCat, into = c("temp1","temp2"), sep=", ", fill = "right") %>% # split rows with multiple functCats
  gather(key = "temp", value = "functCat", temp1, temp2) %>%
  select(-temp) %>%
  drop_na(functCat) %>%
  select(orthogroup, functCat, cladeNumber_split) %>%
  distinct() %>%
  group_by(functCat, cladeNumber_split) %>%
  summarise(n = n()) %>%
  left_join(eggnogFunctCat) %>%
  mutate(functCatDescription = str_c(functCat, ": ", description)) %>%
  ggplot(aes(x = reorder(functCatDescription, -n), y = n)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  ylab("Orthogroup count") +
  xlab("") +
  facet_wrap(~cladeNumber_split, scales = "free_x")
```

# For clade1 only

```{r}
clade1_unique <- gene_pres_abs %>%
  filter(ngenes > 0) %>%
  select(orthogroup, cladeNumber_split) %>%
  distinct() %>%
  add_count(orthogroup) %>%
  filter(n == 1,
         cladeNumber_split == "clade1") %>%
  select(-n) %>%
  left_join(eggnog %>% select(orthogroup, functCat, `eggNOG annot`))
```

349 unique orthogroups for clade 1

```{r, fig.width=8}
clade1_unique %>%
  mutate(functCat = ifelse(is.na(functCat), "S", functCat)) %>% # if function is unkown add to category S
  separate(functCat, into = c("temp1","temp2"), sep=", ", fill = "right") %>% # split rows with multiple functCats
  gather(key = "temp", value = "functCat", temp1, temp2) %>%
  select(-temp) %>%
  drop_na(functCat) %>%
  select(orthogroup, functCat)%>%
  distinct() %>%
  group_by(functCat) %>%
  summarise(n = n()) %>%
  left_join(eggnogFunctCat) %>%
  mutate(functCatDescription = str_c(functCat, ": ", description)) %>%
  ggplot(aes(x = reorder(functCatDescription, -n), y = n)) +
  geom_col(fill = "#1f77b4") +
  geom_text(aes(y = -20, label = n)) +
  coord_flip() +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text.y = element_text(face = "bold", size = 10.2)) +
  ylab("Orthogroup count") +
  xlab("")
```

Let's dig a little bit further into these uniques

```{r echo=TRUE, message=FALSE, warning=FALSE}
# By S. Wittouck
tidy_orthofinder <- function(orthofinder_dir) {
  
  genes_assigned <- orthofinder_dir %>%
    paste0("/Orthogroups.csv") %>%
    readr::read_tsv(col_names = T) %>%
    rename(orthogroup = X1) %>%
    gather(key = "genome", value = "gene", na.rm = T, - orthogroup) %>%
    separate_rows(gene, sep = ", ")
  
}

# Read in gene info
files <- str_c("../11_gene_info/out/", list.files("../11_gene_info/out/",  pattern = "infoseq")) 
names(files) <- files

genes <- map_df(files, read_tsv, skip = 1, col_names = c("gene", "Length", "%GC", "Description"), .id = "filename")  %>%
  separate(filename, "/", into = c("deleteme", "deleteme2", "deleteme3","file"), fill = "right") %>%
  separate(file, "\\.", into = c("genome", "deleteme4"), fill = "right") %>%
  select(-(starts_with("deleteme")))

# Read in orthogroup info
orthogroups <- tidy_orthofinder("~/serverUA/sander/mudan/05_pangenome/Results_Sep03/")

# Merge with uniques

clade1_unique_annot <- clade1_unique %>%
  mutate(functCat = ifelse(is.na(functCat), "S", functCat)) %>% # if function is unkown add to category S
  separate(functCat, into = c("temp1","temp2"), sep=", ", fill = "right") %>% # split rows with multiple functCats
  gather(key = "temp", value = "functCat", temp1, temp2) %>%
  select(-temp) %>%
  drop_na(functCat) %>%
  select(orthogroup, functCat)%>%
  distinct() %>%
  left_join(eggnogFunctCat) %>%
  left_join(orthogroups) %>%
  left_join(genes)
```

Unique annotations for category G

```{r}
clade1_unique_annot %>%
  filter(functCat == "G") %>%
  pull(Description) %>%
  unique() %>%
  kable()
```

Endoglucanase E1

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3856469/:
Conversion of cellulose polymers to simple sugars requires the use of cellulases. Endoglucanases act by cleaving internal β-glycosidic bonds in the cellulose chain, thereby making chain ends accessible to cellobiohydrolase. Endoglucanases have an enzyme classification number 3.2.1.4 and belong to the broader enzyme group called glycosyl hydrolases, which also includes other cellulases such as exoglucanase and β-glucosidase. According to the CAZy database (http://www.cazy.org), endoglucanases are part of 13 distinct glycosyl hydrolase families, distributed in several archeal, bacterial, fungal, and eukaryotic organisms.

Vergelijken met onze eigen CAZYME classificatie! In AMB-F197 werd dit gen inderdaad al aangeduid als een GH5!

In discussie dieper ingaan op dit cellulose content (zie slack sarah)
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3550877/

Van de rest valt weinig te zeggen

```{r}
clade1_unique_annot %>%
  filter(functCat == "M") %>%
  pull(Description) %>%
  unique() %>%
  kable()
```


# UpsetR

Create an upset plot comparing the number of shared orthogroups per species


```{r, fig.height=8, fig.width = 12}
upsetdf <- gene_pres_abs %>%
  select(orthogroup, cladeNumber_split, ngenes) %>% 
  group_by(orthogroup, cladeNumber_split) %>%
  summarise(ngenes = sum(ngenes)) %>% # Sum the number of orthogroups per clade
  mutate(ngenes = if_else(ngenes > 0, 1, 0)) %>% # Convert to presence absence instead of gene count
  mutate(cladeNumber_split = case_when(cladeNumber_split == "clade1" ~ "L.mudanjiangensis",
                               cladeNumber_split == "clade2" ~ "L.fabifermentans",
                               cladeNumber_split == "clade3" ~ "L.xiangfangensis",
                               cladeNumber_split == "clade4" ~ "L.herbarum",
                               cladeNumber_split == "clade5a" ~ "clade5a",
                               cladeNumber_split == "clade5b" ~ "L.pentosus",
                               cladeNumber_split == "clade6" ~ "L.paraplantarum",
                               cladeNumber_split == "clade7" ~ "L.plantarum")) %>%
  spread(cladeNumber_split, ngenes) # Make it untidy again. What a pitty
  
pdf("upset.pdf", width = 12, height = 8)

upsetdf <- as.data.frame(upsetdf)
upset(upsetdf,
      sets = c("L.mudanjiangensis",
               "L.fabifermentans",
               "L.xiangfangensis",
               "L.herbarum",
               "clade5a",
               "L.pentosus",
               "L.paraplantarum",
               "L.plantarum"),
      empty.intersections = "on",
      keep.order = T,
       queries = list(
         list(query = intersects, params = list("L.mudanjiangensis"), active = T),
         list(query = intersects, params = list("L.mudanjiangensis", "L.fabifermentans"), active = T, color = "orange"),
         list(query = intersects, params = list("L.mudanjiangensis", "L.xiangfangensis"), active = T, color = "orange"),
         list(query = intersects, params = list("L.mudanjiangensis", "L.herbarum"), active = T, color = "orange"),
         list(query = intersects, params = list("L.mudanjiangensis", "clade5a"), active = T, color = "orange"),
         list(query = intersects, params = list("L.mudanjiangensis", "L.pentosus"), active = T, color = "orange"),
         list(query = intersects, params = list("L.mudanjiangensis", "L.paraplantarum"), active = T, color = "orange"),
         list(query = intersects, params = list("L.mudanjiangensis", "L.plantarum"), active = T, color = "orange"),
         list(query = intersects, params = list("L.mudanjiangensis", "L.pentosus","L.plantarum"), active = T, color = "red"),
         list(query = intersects, params = list("L.mudanjiangensis", "L.paraplantarum","L.plantarum"), active = T, color = "red")
         ),
      mainbar.y.label = "Number of orthogroups",
      sets.x.label = "Total number of orthogroups",
      point.size = 3.5,
      line.size = 2,
      text.scale = c(2, 1.5, 1.5, 1.5, 2, 1)
      )

dev.off()

```



# Mudan only

```{r, fig.height=8, fig.width = 12}
upsetdf <- gene_pres_abs %>%
  filter(cladeNumber == "clade1") %>%
  select(orthogroup, genome, ngenes) %>% 
  mutate(ngenes = if_else(ngenes > 0, 1, 0)) %>% # Convert to presence absence instead of gene count
  spread(genome, ngenes) # Make it untidy again. What a pitty
  
upsetdf <- as.data.frame(upsetdf)
upset(upsetdf,
      empty.intersections = "on",
      mainbar.y.label = "Number of unique orthogroups",
      sets.x.label = "Total number of orthogroups",
      point.size = 3.5,
      line.size = 2,
      text.scale = c(2, 1.5, 1.5, 1.5, 2, 1),
      order.by = "freq"
      )
```


